sdmTMB: Fast, flexible, and user-friendly spatial and spatiotemporal generalized linear mixed-effects models

### 2021-12-29 doc/paper.Rmd
minor edits to validation and package comparison

Validation and selection of state-space models is challenging, particularly when using the Laplace approximation [@thygesen2017a]. 
We provide several approaches to assist this process: 
(1) The Akaike Information Criterion [AIC, @akaike1974] can be calculated with `AIC()`, although AIC has well-documented biases with mixed-effects models [@liang2008].
<!-- We allow AIC to be calculated for each sdmTMB model [`AIC()`, @akaike1974]; however, because of well-documented biases of AIC with mixed-effects models [@liang2008] we provide functionality to perform k-fold cross-validation. -->
<!-- The function `sdmTMB_cv()` allows for any number of folds, with assignments being user-specified or chosen randomly. -->
<!-- The predictive log-density is calculated for each fold and summarized across folds for a total measure of the predictive accuracy of a model. -->
<!-- Validation of state-space models is challenging, particularly when using the Laplace approximation [@thygesen2017a]. We provide four approaches to assist with model validation: -->
(2) Alternatively, k-fold cross validation with `sdmTMB_cv()` can be used with user-specified or randomly chosen folds for model selection [e.g., via expected log predictive density [ELPD], @vehtari2017] or to evaluate goodness of fit according to user-calculated criteria (e.g., mean squared error, area under the curve).
(3) An sdmTMB model can be passed to the tmbstan package [@monnahan2018] to sample from the joint posterior with Stan [@carpenter2017], evaluate the accuracy of the Laplace approximation, or evaluate residuals (see `?extract_mcmc`). 
(4) The `residuals()` method returns randomized quantile [@dunn1996] or probability integral transform (PIT) [@smith1985] residuals.
<!-- These residuals, however, are calculated as $\hat{y} - y$ where $\hat{y}$ is the sum of the fixed effect estimates and the random effect values that maximize the likelihood conditional on the estimated fixed effects.  -->
For state-space models, these residuals have known statistical issues [@thygesen2017a] but are quick to calculate.
Simulation-based residuals [@dharma], one-step-ahead residuals [@thygesen2017a], or residuals from a single MCMC draw [e.g., @rufener2021] are also possible, but are slower.
(5) The `simulate.sdmTMB()` method or `sdmTMB_simulate()` function can simulate from fitted models or new data to which models can be fit to ensure identifiability, evaluate bias and precision in parameter estimation, or evaluate the consequences of model misspecification.
<!-- Simulations can also be passed to the R package DHARMa [@DHARMa] to calculate simulation-based residual diagnostics (APPENDIX OR VIGNETTE). -->
<!-- We note this will be considerably slower to fit than with maximum likelihood and in some cases may require tight priors or fixed parameters to ensure adequate sampling [as in @monnahan2021]. -->

@@

There are many R packages capable of fitting geostatistical spatial or spatiotemporal models [e.g., @heaton2017].
sdmTMB, VAST, INLA/inlabru, and spaMM [@rousset2014] are the most closely related, as they all provide a user interface to SPDE-based GMRF models; we also include spBayes as a prominent package that can fit related predictive-process models without the SPDE (Table 1).
sdmTMB, VAST, and mgcv can estimate anisotropic covariance whereas INLA/inlabru and spBayes are currently limited to isotropic covariance.
sdmTMB and mgcv focus on univariate response data, whereas VAST, INLA/inlabru, spaMM, and spBayes extend to multivariate responses with varying limitations.
To our knowledge, VAST is the only package to implement spatial [@thorson2015] and spatial dynamic factor analysis [@thorson2016]. 
<!-- However, VAST is typically less effective at replicating nonlinear preferences than GAMs, which are supported by several of these packages [@brodie2020]. -->
<!-- and spatial empirical orthogonal function (EOF) regression [@thorson2020]. To my reading EOF was not implemented in VAST. It seemed to me its own thing?-->
Of these packages, only sdmTMB and inlabru can currently fit threshold (e.g., hockey-stick) covariate relationships.
spaMM is limited to a spatial random field and spBayes implements spatiotemporal fields, but only as a random walk.
<!-- TODO: still true? -->
There is considerable variability in the available observation likelihoods across packages (Table 1).
<!-- TODO: add something about the results of the INLA comparison (Appendix 4) and the VAST comparison (Appendix 5) -->
We provide comparisons of the syntax and the reproducibility of results from models fit using INLA (Appendix 4) and VAST (Appendix 5).
<!-- We ran into problems with AR1 spatiotemporal fields in inlabru, otherwise we would have included both INLA and inlabru syntax.
<!-- TODO: mention inlabru here -->

@@

We ran a simple speed comparison between sdmTMB, inlabru, and mgcv for fitting a SPDE spatial random field model to 1000 data points with Gaussian error across a range of mesh resolutions (Fig. \@ref(fig:timing), Appendix 6).
In this test, for mesh resolutions under 500 nodes, sdmTMB fit the models the fastest.
inlabru was the least affected by mesh resolution and beyond 500 nodes was faster than sdmTMB with default sdmTMB settings.
mgcv was slightly slower than sdmTMB and was more affected by mesh resolution.
With the random effect normalizing constant calculated in R (`normalize = TRUE`), sdmTMB was the fastest across all mesh resolutions (this option may not be possible or as fast for all models).
Results with up to an order of magnitude more data slightly shift the lines up without affecting the overall pattern substantially (not shown).
Our test was with one core and default R algebra libraries; all packages could run faster with optimized libraries and parallel processing.
Results with optimized math libraries (openBLAS: @penblas2021 and PARDISO: @pardiso-7.2c) resulted in a ~10% speed increase for sdmTMB and inlabru and a ~7--9-fold speed increase for mgcv.
<!-- Note two other references for PARDISO were listed as required here https://pardiso-project.org/manual/manual.pdf -->

>>>>

Validation and selection of state-space models is challenging, particularly when using the Laplace approximation [@thygesen2017a]. 
We provide several approaches to assist this process: 
(1) The Akaike Information Criterion [AIC, @akaike1974] can be calculated with `AIC()`, although AIC has well-documented biases with mixed-effects models [@liang2008].
<!-- We allow AIC to be calculated for each sdmTMB model [`AIC()`, @akaike1974]; however, because of well-documented biases of AIC with mixed-effects models [@liang2008] we provide functionality to perform k-fold cross-validation. -->
<!-- The function `sdmTMB_cv()` allows for any number of folds, with assignments being user-specified or chosen randomly. -->
<!-- The predictive log-density is calculated for each fold and summarized across folds for a total measure of the predictive accuracy of a model. -->
<!-- Validation of state-space models is challenging, particularly when using the Laplace approximation [@thygesen2017a]. We provide four approaches to assist with model validation: -->
(2) Alternatively, k-fold cross validation with `sdmTMB_cv()` can be used with user-specified or randomly chosen folds for model selection [e.g., via expected log predictive density [ELPD], @vehtari2017] or to evaluate goodness of fit according to user-calculated criteria (e.g., mean squared error, area under the curve).
(3) A sdmTMB model can be passed to the tmbstan package [@monnahan2018] to sample from the joint posterior with Stan [@carpenter2017], evaluate the accuracy of the Laplace approximation, or evaluate residuals (see `?extract_mcmc`). 
(4) The `residuals()` method returns randomized quantile [@dunn1996] or probability integral transform (PIT) [@smith1985] residuals.
<!-- These residuals, however, are calculated as $\hat{y} - y$ where $\hat{y}$ is the sum of the fixed effect estimates and the random effect values that maximize the likelihood conditional on the estimated fixed effects.  -->
For state-space models, these residuals have known statistical issues [@thygesen2017a] but are quick to calculate.
Simulation-based residuals [@dharma], one-step-ahead residuals [@thygesen2017a], or residuals from a single MCMC draw [e.g., @rufener2021] are also possible, but are slower.
(5) The `simulate.sdmTMB()` method or `sdmTMB_simulate()` function can simulate from fitted models or new data to which models can be fit to ensure identifiability, evaluate bias and precision in parameter estimation, or evaluate the consequences of model misspecification.
<!-- Simulations can also be passed to the R package DHARMa [@DHARMa] to calculate simulation-based residual diagnostics (APPENDIX OR VIGNETTE). -->
<!-- We note this will be considerably slower to fit than with maximum likelihood and in some cases may require tight priors or fixed parameters to ensure adequate sampling [as in @monnahan2021]. -->

@@

There are many R packages capable of fitting geostatistical spatial or spatiotemporal models [e.g., @heaton2017].
sdmTMB, VAST, INLA/inlabru, and spaMM [@rousset2014] are the most closely related, as they all provide a user interface to SPDE-based GMRF models; we also include spBayes in our software comparison as it is a prominent package that can fit related predictive-process models without the SPDE (Table 1).
sdmTMB, VAST, and mgcv can estimate anisotropic covariance whereas INLA/inlabru and spBayes are currently limited to isotropic covariance.
sdmTMB and mgcv focus on univariate response data, whereas VAST, INLA/inlabru, spaMM, and spBayes extend to multivariate responses with varying limitations.
To our knowledge, VAST is the only package to implement spatial [@thorson2015] and spatial dynamic factor analysis [@thorson2016]. 
<!-- However, VAST is typically less effective at replicating nonlinear preferences than GAMs, which are supported by several of these packages [@brodie2020]. -->
<!-- and spatial empirical orthogonal function (EOF) regression [@thorson2020]. To my reading EOF was not implemented in VAST. It seemed to me its own thing?-->
Of these packages, only sdmTMB and inlabru can currently fit threshold (e.g., hockey-stick) covariate relationships.
spaMM is limited to a spatial random field and spBayes implements spatiotemporal fields, but only as a random walk.
<!-- TODO: still true? -->
There is considerable variability in the available observation likelihoods across packages (Table 1).
<!-- TODO: add something about the results of the INLA comparison (Appendix 4) and the VAST comparison (Appendix 5) -->
We provide comparisons of the syntax and the reproducibility of results from models fit using INLA (Appendix 4) and VAST (Appendix 5).
<!-- We ran into problems with AR1 spatiotemporal fields in inlabru, otherwise we would have included both INLA and inlabru syntax.
<!-- TODO: mention inlabru here -->

@@

We ran a simple speed comparison between sdmTMB, inlabru, and mgcv for fitting a SPDE spatial random field model to 1000 data points with Gaussian error across a range of mesh resolutions (Fig. \@ref(fig:timing), Appendix 6).
In this test, for mesh resolutions under 500 nodes, sdmTMB fit the models the fastest.
inlabru was the least affected by mesh resolution and beyond 500 nodes was faster than sdmTMB with default sdmTMB settings.
mgcv was slightly slower than sdmTMB and was more affected by mesh resolution.
With the random effect normalizing constant calculated in R (`normalize = TRUE`), sdmTMB was the fastest across all mesh resolutions (this option may not be possible or as fast for all models).
Results with up to an order of magnitude more data slightly shift the lines up without affecting the qualitative pattern (not shown).
Our test was with one core and default R algebra libraries; all packages could run faster with optimized libraries and parallel processing.
Results with optimized math libraries (openBLAS: @penblas2021 and PARDISO: @pardiso-7.2c) resulted in a ~10% speed increase for sdmTMB and inlabru and a ~7--9-fold speed increase for mgcv.
<!-- Note two other references for PARDISO were listed as required here https://pardiso-project.org/manual/manual.pdf -->
<!-- LB: also worth noting which packages do/do not have parallel infrastructure (notably that VAST does not) -->
